Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-30 paper/gd1.tex
Orbit text

To study the global properties of the stream as a function of stream longitude,
we extract stream stars around a low-order polynomial model for the median
stream latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
We find stream ridge points by computing a running median in $\phi_2$ in steps
of $2^\circ$ with a window size of $4^\circ$ along $\phi_1$, then fit a
second-order polynomial to the ridge points to find $\phi_{2,
\textrm{track}}(\phi_1)$.
We then define the stream as the region within $\left| \phi_2 - \phi_{2,
\textrm{track}}(\phi_1) \right| < 0.75^\circ$.
\figurename~\ref{fig:track-and-model} (top panel) again shows the
high-confidence stream stars, and the two curved lines (blue) show the adopted
upper and lower boundary of the stream region.

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing median stream properties in overlapping windows:
we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ in longitude for each subsequent measurement.
Panels 3--5 in \figurename~\ref{fig:track-and-model} show the
background-subtracted stream surface density and median proper motion
components, $\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way (dark
lines), with the associated empirical scatters (shaded regions).
We estimate the local background surface density as an average of areas north
and south of the bounded stream region.

@@

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These are not a result of variable dust extinction or completeness variations
from the \gaia\ scanning pattern.
First, the visible portions of the GD-1 stream are located at high Galactic
latitudes ($b > 20^\circ$), and we therefore do not expect significant dust
extinction or variations in extinction within the stream region.
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
\approx 0.03~\textrm{mag}$, as computed from the SFD
extinction map (\citealt{Schlegel:1998}).
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable, and the stream stars are harder to
select apart from the background.
Second, each star in our sample has at least XX visibility periods, while the sample median and variance are XX and XX, respectively, guaranteeing a robust estimate of proper motions.
Variations in the number density of faint stars originating from the scanning pattern are XX times smaller than the density variations in the stream, and their spatial distribution does not map to the stream features.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
\todo{check above sentences +  fill in}

The density variations likely represent real morphological changes along the
GD-1 stream, as have been suggested before from binned samples of
color-magnitude-selected GD-1 stars (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp over-density in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
% \todo{Table of locations, sizes, and density constrasts for the features?}

We compare the measured stream properties to a simple model for the phase-space
density of the stream, generated by only simulating the orbital evolution of
cluster stars once they are tidally stripped from the progenitor.
In detail, during simulation, star particles are released from the Lagrange
points of the cluster with a spread in position and velocity set by the mass of
the satellite and the location within the host galaxy, with some tunable scale
parameters.
We use the parametrization and adopted scale parameters of \citet{Fardal:2015},
who fit the release distribution parameters by comparing to full $N$-body
simulations across a range of mass and orbital parameters.
This method can accurately reproduce the mean track of the resulting stream, but
the density of stars along the stream depends strongly on the mass-loss history
and internal kinematics of the progenitor system.
Here we assume a constant mass-loss rate, and release stream particles uniformly
in time from each Lagrange point; these choices match results from more detailed numerical experiments of globular cluster disruption \citep[e.g.,][]{Kupper:2012}.

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed model for the gravitational
potential of the Milky Way.
We use a three-component potential model to represent the Galaxy, consisting of
a disk (\citealt{Miyamoto:1975}), a bulge
(\citealt{Hernquist:1990}), and a spherical dark matter halo
(\citealt{Navarro:1996}).
We set the disk scale-length parameters to closely match the disk model found in
\citet{Bovy:2015}, and other parameters are set to reproduce an approximately
flat rotation curve between $5 < R < 15~\kpc$ with a circular velocity at the
solar circle, $v_{\textrm{circ}, \odot} \approx 230~\kms$.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{orbitfit.pdf}
\end{center}
\caption{%
\todo{APW}
}
\label{fig:orbitfit}
\end{figure}

To initialize the model stream generation, we assume that the present day (i.e.
final) position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak
of the observed surface density profile.
We integrate backwards for $4~\textrm{Gyr}$ using the orbital parameters at this
location using the Leapfrog integration scheme with a timestep of
$0.5~\textrm{Myr}$, then begin the model stream generation using the Lagrange
point stripping method discussed above.
Stream particles are release at each timestep of the progenitor orbit
integration, resulting in 16,000 stream particles by the end of the model stream
generation.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until present day, where we assume $M = 0~\msun$
(i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as \todo{XX}.

\figurename~\ref{fig:orbitfit} shows \todo{APW: ...}

>>>>

To study the global properties of the stream as a function of stream longitude,
we extract stream stars around a low-order polynomial model for the median
stream latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
We find stream ridge points by computing a running median in $\phi_2$ in steps
of $2^\circ$ with a window size of $4^\circ$ along $\phi_1$, then fit a
second-order polynomial to the ridge points to find $\phi_{2,
\textrm{track}}(\phi_1)$.
We then define the stream as the region within $\left| \phi_2 - \phi_{2,
\textrm{track}}(\phi_1) \right| < 0.75^\circ$.
\figurename~\ref{fig:track-and-model}, panel (a), shows the high-confidence
stream stars (same as in lower panel of \figurename~\ref{fig:selection}), and
the two curved lines (blue) show the adopted upper and lower boundary of the
stream region.

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing median stream properties in overlapping windows:
we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ in longitude for each subsequent measurement.
Panels (c)--(e) in \figurename~\ref{fig:track-and-model} show the
background-subtracted stream surface density and median proper motion
components, $\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way (dark
lines), with the associated empirical scatters (shaded regions).
We estimate the local background surface density as an average of areas north
and south of the bounded stream region.

@@

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These are not a result of variable dust extinction or completeness variations
from the \gaia\ scanning pattern.
First, the visible portions of the GD-1 stream are located at high Galactic
latitudes ($b > 20^\circ$), and we therefore do not expect significant dust
extinction or variations in extinction within the stream region.
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
\approx 0.03~\textrm{mag}$, as computed from the SFD
extinction map (\citealt{Schlegel:1998}).
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable, and the stream stars are harder to
select apart from the background.
Second, 99\% of the stars in our sample have at least 8 visibility periods,
while the sample median and root-variance in number of visibility periods are 14
and 3, respectively, guaranteeing a robust estimate of proper motions.
The variance in the number density of faint stars originating from the \gaia\
scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
\mu_{\phi_2, \odot} < 4~\masyr$ (i.e. above the stream selection box), far
smaller than the density variations in the stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they

The density variations likely represent real morphological changes along the
GD-1 stream, as have been suggested before from binned samples of
color-magnitude-selected GD-1 stars (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp over-density in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material that has not yet begun to phase-mix in multiple
dimensions.

We also measured the median proper motion profile along the stream (panels
(d)--(e)) and find that these are consistent with previously measured mean
proper motions from SDSS-USNOB (\citealt{Koposov:2010}).
Here, the proper motion gradient is resolved in striking detail for individual
stars from \gaia\ (see, e.g., elongated over-density in upper right panel of
\figurename~\ref{fig:selection}).

We compare the measured stream properties to a simple model for the phase-space
density of the stream, generated by only simulating the orbital evolution of
cluster stars once they are tidally stripped from the progenitor.
During simulation, star particles are released from the Lagrange points of the
cluster with a spread in position and velocity set by the mass of the satellite
and the location within the host galaxy, with some tunable scale parameters
(using the parametrization and adopted scale parameters of \cite{Fardal:2015}).
This method can accurately reproduce the mean track of the resulting stream, but
the density of stars along the stream depends in detail on the mass-loss history
and internal kinematics of the progenitor system.
Here we assume a constant mass-loss rate, and release stream particles uniformly
in time from each Lagrange point; these choices match results from more detailed
numerical experiments of globular cluster disruption
\citep[e.g.,][]{Kupper:2012}.

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed model for the gravitational
potential of the Milky Way.
We use a three-component potential model to represent the Galaxy, consisting of
a disk (\citealt{Miyamoto:1975}), a bulge
(\citealt{Hernquist:1990}), and a spherical dark matter halo
(\citealt{Navarro:1996}).
We set the disk scale-length parameters to closely match the disk model found in
\citet{Bovy:2015}, and other parameters are set to reproduce an approximately
flat rotation curve between $5 < R < 15~\kpc$ with a circular velocity at the
solar circle, $v_{\textrm{circ}, \odot} \approx 230~\kms$.
We use L-BFGS-B optimization as implemented in the \package{scipy}
\texttt{Python} package to maximize the likelihood of the orbit given sky track
and proper motions from this work with mean distance and giant star radial
velocities from \cite{Koposov:2010}.
We do not vary the parameters of the Milky Way model in the fit because we are
only interested in producing a comparison model of the stream properties.

\figurename~\ref{fig:orbitfit} shows the best-fitting orbit from the procedure
outlined above.
The GD-1 stream crosses the midplane at large radius, far from the bulk of the
stellar density:
the grey line in the left panel shows the radius that contains $\approx 90\%$ of
the mass, assuming an exponential radial density profile with a scale radius
$\R_d = 3~\kpc$ (an intermediate value of recent measurements, e.g.,
\citealt{Bovy:2012}).

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{orbitfit.pdf}
\end{center}
\caption{%
Best-fitting orbit in a simple (fixed) model for the gravitational potential of
the Milky Way using phase-space information along the stream measured from this
work combined with data from \citet{Koposov:2010}.
Left panel shows the orbit in Galactocentric cylindrical radius, $R$, and height
above the Galactic midplane, $z$;
grey line at $z=0$ shows the approximate radius of the disk.
Right panel shows the orbit projected onto the midplane in Galactocentric
cartesian coordinates.
The darker orbit section highlights the range of stream longitudes over which we
clearly see GD-1 stream stars in \figurename~\ref{fig:selection}.
Note that GD-1 is retrograde with respect to the disk (see approximate location
and direction of motion of the sun in right panel, $\odot$).
}
\label{fig:orbitfit}
\end{figure}

To initialize the model stream generation, we assume that the present day (i.e.
final) position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak
of the observed surface density profile.
We integrate backwards for $4~\textrm{Gyr}$ using the orbital parameters at this
location using the Leapfrog integration scheme with a timestep of
$0.5~\textrm{Myr}$, then forward-generate the model stream using the method
discussed above.
Stream particles are release at each timestep of the integrated progenitor
orbit.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until present day, where we assume $M = 0~\msun$
(i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
